---
title: "Exploratory & Confirmatory Analysis"
author: "LIN XIANWEI"
date: today
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  warning: false
  message: false
---

# Storyboard Baseline

Through a rigorous pipeline of Exploratory Data Analysis (EDA), Confirmatory Data Analysis (CDA), and advanced Machine Learning techniques, we have successfully transformed raw transactional logs into a predictive retention engine.

### Key Analytical Takeaways:

1.  **The Friction Point is Proven:** Our CDA (`ggscatterstats` and Correlation Matrix) mathematically confirmed that customer support bottlenecks are the primary driver of negative sentiment and churn risk, even among high-volume users.
2.  **The 90-Day Danger Zone:** The Kaplan-Meier Survival Analysis revealed a critical drop-off in customer retention between Month 3 and Month 6, highlighting the exact time window where marketing interventions are most necessary.
3.  **Liquidity is Tied to Retention:** The Sankey diagram illustrated the macro-financial impact of churn. When users leave, capital flows directly out of the app wallet to external accounts, causing a massive bleed in platform liquidity.
4.  **Predictive Interventions Work:** The Multiple Linear Regression model proves that churn is not random—it is highly sensitive to actionable variables like `support_tickets_count` and `app_logins_frequency`.

### Next Steps: The Interactive Dashboard

Static analysis provides the "what" and the "why," but decision-makers need tools to simulate the "what if." All of the statistical models, clustering algorithms (K-Means), and datasets generated in this document have been integrated into the **FinRetain Shiny Application**. Stakeholders can now use the dashboard's **Risk Simulator** to dynamically adjust user parameters and visualize the real-time impact on Customer Lifetime Value (CLV) and Churn Probability.

## 1. Loading packages

```{r}
# Load all required libraries for advanced interactive analytics
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(survival)
library(networkD3)
library(scales)
library(broom) # For tidying survival models

# Load the core application dataset
df <- readRDS("app_data.rds")

# Define FinTech Color Palette
fintech_colors <- c("#1c2541", "#0077b6", "#00b4d8", "#90e0ef", "#caf0f8")
risk_color <- "#e74c3c"

# Load the clean data generated from our preparation pipeline
clean_customers <- readRDS("app_data.rds")

# Prepare churn as a factor for categorical plotting
clean_customers$churn_factor <- factor(clean_customers$churn_event, 
                                       levels = c(0, 1), 
                                       labels = c("Active", "Churned"))
```

## 2. Demographics (EDA)

Our EDA phase focuses on understanding *who* our customers are. Using `ggplot2` and `plotly` in our application, stakeholders can dynamically filter these demographic distributions by gender and customer segment.

### 2.1 Joy / Ridgeline Plots for Distributions - Tenure Distribution by Segment

How does Customer Tenure (how long they stay) distribute across different RFM Segments? Are "Champions" naturally older accounts, while "At Risk" users are newer? By using a ridgeline plot, we visualize the density of customer tenure across our RFM Segments. This helps us see if our "Champions" are primarily older accounts, or if recent users are also reaching champion status.

```{r}
#| label: eda-ridgeline
#| warning: false
#| message: false

# Install ggridges if you haven't: install.packages("ggridges")
library(ggplot2)
library(ggridges)

ggplot(clean_customers, aes(x = customer_tenure, y = RFM_Segment, fill = RFM_Segment)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5) +
  scale_fill_viridis_d(option = "mako") +  # Professional color palette
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Customer Tenure Distribution by RFM Segment",
    x = "Tenure (Months)",
    y = "RFM Segment"
  )
```

### 2.2 Income and Education Distributions (EDA)

Our EDA phase focuses on understanding *who* our customers are. We visualized the distribution of customers across education levels to see how age correlates with educational attainment.

```{r}
#| fig-align: "center"

p_eda <- ggplot(df, aes(x = education_level, y = age, fill = education_level)) +
  geom_boxplot(alpha = 0.8, color = "#1c2541", outlier.color = risk_color) +
  scale_fill_manual(values = colorRampPalette(fintech_colors)(length(unique(df$education_level)))) +
  theme_minimal() +
  labs(title = "Age Distribution by Education Level", 
       x = "Education Level", 
       y = "Age") +
  theme(legend.position = "none", 
        panel.grid.minor = element_blank())

# Convert to interactive Plotly object
ggplotly(p_eda) %>% config(displayModeBar = FALSE)
```

## 3. Behavioral Analysis (CDA)

The CDA phase seeks to test hypotheses regarding how customer *behavior* impacts overall business metrics like Customer Lifetime Value (CLV) and Net Promoter Score (NPS).

### 3.1 Correlation Matrix

Before diving into predictive modeling, we use a correlation matrix to identify the strength and direction of relationships between our numerical variables. This helps us identify "multicollinearity" and select the most impactful features for our churn prediction model.

```{r}
# 3. Statistical Correlation Analysis
#| label: fig-correlation
#| fig-cap: "Correlation Matrix of Behavioral and Risk Metrics"
#| warning: false
#| message: false

library(ggcorrplot)
library(dplyr)

# 1. Load the processed data
df <- readRDS("app_data.rds")

# 2. Select numerical columns relevant to our business questions
corr_data <- df %>%
  select(
    `Age` = age, 
    `Logins` = app_logins_frequency, 
    `Tx Count` = tx_count, 
    `Volume` = total_tx_volume, 
    `Satisfaction` = satisfaction_score, 
    `Tickets` = support_tickets_count, 
    `Tenure` = customer_tenure, 
    `Churn Risk` = churn_probability
  ) %>%
  # Remove rows with missing values to ensure accurate calculation
  filter(complete.cases(.))

# 3. Compute the correlation matrix
corr_matrix <- cor(corr_data)

# 4. Render the Heatmap
ggcorrplot(corr_matrix, 
           hc.order = TRUE,           # Clusters similar variables together
           type = "lower",            # Shows only the bottom half (cleaner look)
           outline.col = "white",
           lab = TRUE,                # Shows the correlation coefficients (r)
           lab_size = 3.5,
           colors = c("#e46764", "white", "#6495ed"), # Red (Negative) to Blue (Positive)
           title = "Behavioral Correlation Matrix",
           legend.title = "Correlation (r)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

### 3.2 Custom Correlation Function: Transactions vs. Satisfaction

To ensure scalability, we built a custom wrapper function to test the nonparametric (Spearman) correlation between Transaction Count and Satisfaction Score, allowing us to dynamically facet by any demographic variable.

```{r}
clean_customers <- readRDS("app_data.rds") 

# 2. Load the required libraries
library(ggstatsplot)
library(rlang)

# 3. Define our custom FinTech correlation function
plotcorrelation_tx_v_sat <- function(facet_var) {
  ggscatterstats(
    data = clean_customers,
    x = tx_count, 
    y = satisfaction_score,
    type = "nonparametric", 
    title = paste("Transaction Volume vs Satisfaction faceted by", facet_var),
    messages = FALSE
  ) + 
  facet_wrap(vars(!!sym(facet_var)))
}

# 4. Call the function
plotcorrelation_tx_v_sat("income_bracket")
```

### 3.3 Group Comparisons with `ggbetweenstats`: Volume across Education Levels (CDA)

Does the total transaction volume differ significantly across different education levels? Does a user's education level statistically impact the total monetary volume they push through the app? We use the Kruskal-Wallis test to confirm.

```{r}
ggbetweenstats(
  data = clean_customers,
  x = education_level,
  y = total_tx_volume,
  type = "nonparametric",
  title = "Transaction Volume Distribution by Education Level",
  ylab = "Total Transaction Volume ($)",
  xlab = "Education Level",
  messages = FALSE
)
```

### 3.4 Proportional Testing: Churn vs. Income - Proportions & Chi-Square with `ggbarstats` (CDA)

Finally, we test categorical dependency. Is the proportion of churned users statistically dependent on their Income Bracket? The Chi-square test automatically calculates the p-value for this relationship.

```{r}
#| label: cda-barstats
#| warning: false
#| message: false

library(ggstatsplot)

# We need churn_event as a factor for this plot
clean_customers$churn_factor <- factor(clean_customers$churn_event, 
                                       levels = c(0, 1), 
                                       labels = c("Active", "Churned"))

# Compare Churn Status against Income Bracket
ggbarstats(
  data = clean_customers,
  x = churn_factor,
  y = income_bracket,
  title = "Churn Proportion across Income Brackets",
  messages = FALSE
)
```

### 3.5 Interactive Correlation Analysis

We deployed interactive scatter plots that allow users to map specific X and Y variables against each other (e.g., *Total Transactions vs. Satisfaction Score*). By overlaying a statistical trendline (using the `lm` method), stakeholders can immediately confirm or reject suspected correlations.

```{r}
#| fig-align: "center"

ggplot(df, aes(x = tx_count, y = satisfaction_score)) +
  geom_point(alpha = 0.5, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e74c3c", se = FALSE) +
  theme_minimal() +
  labs(title = "Total Transactions vs Satisfaction Score",
       x = "Total Transaction Count",
       y = "Satisfaction Score (1-5)")
```

### 3.6 Transaction vs. Satisfaction Hypothesis

We plotted transaction counts against satisfaction scores. The red trendline highlights a concerning friction point: higher transaction volumes do not strictly guarantee higher satisfaction.

```{r}
#| fig-align: "center"

p_cda <- ggplot(df, aes(x = tx_count, y = satisfaction_score, text = paste("Support Tickets:", support_tickets_count))) +
  geom_point(alpha = 0.4, color = "#0077b6") +
  geom_smooth(method = "lm", color = risk_color, se = TRUE, fill = "#ffcccb", alpha = 0.3) +
  theme_minimal() +
  labs(title = "Hypothesis Testing: Usage vs Satisfaction",
       x = "Total Transaction Count",
       y = "Satisfaction Score (1-5)")

ggplotly(p_cda, tooltip = c("x", "y", "text")) %>% config(displayModeBar = FALSE)
```

### 3.7 Transaction Volume Breakdown

We aggregated the total dollar volume flowing through four distinct transactional categories:

1.  **Payments**

2.  **Transfers**

3.  **Withdrawals**

4.  **Deposits**

This breakdown acts as the foundational baseline that later informs the proportional weights used in our **Macro Cash Flow Sankey Diagram**.

```{r}
# Calculate the summaries using the actual columns in the dataset!
library(dplyr)

tx_sums <- df %>%
  group_by(preferred_transaction_type) %>%
  summarise(
    Total_Volume = sum(total_tx_volume, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Volume)) # Sorts it from highest to lowest volume

# Print the final data table to the webpage
tx_sums
```

## 4. Cluster Analysis

### 4.1 K-Means Clustering

To move beyond basic demographic segmentation, we applied a **K-Means Clustering algorithm** (k=3) to automatically group users based on their transaction frequency and monetary volume to discover hidden behavioral groupings within our user base. By scaling and analyzing `tx_count` (Transaction Frequency) against `total_tx_volume` (Monetary Value), the algorithm identified three distinct customer clusters without any manual labeling.

```{r}
#| fig-align: "center"

# 1. Prepare data and apply K-Means
set.seed(42)
cluster_data <- df %>% select(tx_count, total_tx_volume) %>% drop_na()
kmeans_result <- kmeans(scale(cluster_data), centers = 3)
cluster_data$Cluster <- paste("Persona", kmeans_result$cluster)

# 2. Interactive Cluster Plot
p_cluster <- ggplot(cluster_data, aes(x = tx_count, y = total_tx_volume, color = Cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("#0077b6", "#00b4d8", risk_color)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  labs(title = "Algorithmic Behavioral Segmentation",
       x = "Transaction Frequency",
       y = "Total Monetary Volume")

ggplotly(p_cluster) %>% layout(legend = list(orientation = "h", x = 0.3, y = -0.2))
```

### 4.2 RFM Analysis

While K-Means provides algorithmic grouping, **RFM** (Recency, Frequency, Monetary) provides established business logic. We categorized our users based on their \`**R_Score\`, \`F_Score\`, and \`M_Score**\` to identify key groups like "**Champions**" and "**At-Risk**" users. In our Shiny application, this is visualized as an interactive Treemap.

```{r}
## 4. Advanced Segmentation (RFM Analysis)
#| fig-align: "center"
library(dplyr)
library(ggplot2)

# Calculate the size of each RFM Segment
rfm_summary <- df %>%
  group_by(RFM_Segment) %>%
  summarise(Customer_Count = n()) %>%
  arrange(desc(Customer_Count))

# Plotting the Segment Distribution
ggplot(rfm_summary, aes(x = reorder(RFM_Segment, Customer_Count), y = Customer_Count, fill = RFM_Segment)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Customer Distribution by RFM Segment",
       x = "RFM Segment",
       y = "Number of Customers")
```

**Interactive RFM Treemap**

**RFM Analysis** provides established business logic. Here is the entire distribution of our customer base grouped by their Recency, Frequency, and Monetary scores.

```{r}
#| fig-align: "center"

# Calculate RFM Sizes
rfm_summary <- df %>%
  group_by(RFM_Segment) %>%
  summarise(Customer_Count = n()) %>%
  arrange(desc(Customer_Count))

# Build an interactive Plotly Treemap
plot_ly(
  type = "treemap",
  labels = rfm_summary$RFM_Segment,
  parents = rep("All Customers", nrow(rfm_summary)),
  values = rfm_summary$Customer_Count,
  textinfo = "label+value+percent parent",
  marker = list(colorscale = "Blues")
) %>%
  layout(title = "Customer Distribution by RFM Segment")
```

## 5. Transition to Advanced Analytics

With a solid understanding of our demographics and baseline transaction volumes established in this EDA/CDA phase, we confidently move forward to our predictive and advanced modeling phases. The insights gathered here directly feed into the variables chosen for our **Linear Regression Risk Simulator** and **Kaplan-Meier Survival** curves and **Sankey Network**.

### 5.1 Exploratory Modelling

To understand the macro-dynamics of our user base, we explored the dimensions of **Time** (when users leave) and **Capital** (where money flows) by using **Survival Analysis (Time-to-Churn)**. We applied a **Kaplan-Meier Survival Model** to map the probability of a customer remaining active over time. By using `customer_tenure` and the binary `churn_event` indicator, we can pinpoint the exact month where retention efforts must be intensified.

```{r}
#| fig-align: "center"
library(survival)

# Build the Kaplan-Meier survival object
km_fit <- survfit(Surv(customer_tenure, churn_event) ~ 1, data = df)

# Plot the baseline survival curve using base R (survminer is used in the app)
plot(km_fit, 
     xlab = "Customer Tenure (Months)", 
     ylab = "Survival Probability", 
     main = "Kaplan-Meier Estimate of Customer Retention",
     col = "#0077b6", lwd = 2, conf.int = TRUE)
abline(v = 3, col = "red", lty = 2) # Highlight Month 3 risk zone
```

**Interactive Model**

```{r}
#| fig-align: "center"

# Build Kaplan-Meier Model
km_fit <- survfit(Surv(customer_tenure, churn_event) ~ 1, data = df)
km_data <- tidy(km_fit) # Use broom to tidy for ggplot

# Interactive Survival Curve
p_surv <- ggplot(km_data, aes(x = time, y = estimate)) +
  geom_step(color = "#1c2541", size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "#0077b6") +
  geom_vline(xintercept = 3, color = risk_color, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  labs(title = "Kaplan-Meier Retention Curve",
       x = "Customer Tenure (Months)",
       y = "Probability of Retaining User") +
  annotate("text", x = 4.5, y = 0.6, label = "High Risk Zone", color = risk_color)

ggplotly(p_surv) %>% config(displayModeBar = FALSE)
```

### 5.2 Macro Cash Flow (Sankey Network)

To visualize liquidity, we mapped the flow of transactions using our actual transactional dataset. Deposits represent capital entering the FinRetain ecosystem ("**App Wallet**"), while **Payments**, **Transfer**s, and **Withdrawals** represent capital flowing out to various destinations.

```{r}
#| fig-align: "center"
library(readxl)
# 1. Load data & summarize
tx_data <- read_xlsx("transactions_data.xlsx")
tx_summary <- tx_data %>%
  group_by(type) %>%
  summarise(total_volume = sum(amount, na.rm = TRUE))

# 2. Map logical flow
links <- data.frame(
  source = c("External Accounts", "App Wallet", "App Wallet", "App Wallet"),
  target = c("App Wallet", "Merchant Payments", "Peer Transfers", "External Withdrawals"),
  type_name = c("Deposit", "Payment", "Transfer", "Withdrawal")
) %>%
  left_join(tx_summary, by = c("type_name" = "type")) %>%
  rename(value = total_volume) %>% filter(value > 0)

# 3. Calculate Node Totals
node_totals <- data.frame(name = unique(c(links$source, links$target))) %>%
  rowwise() %>%
  mutate(total = max(sum(links$value[links$source == name]), sum(links$value[links$target == name])),
         label = paste0(name, " ($", format(round(total, 0), big.mark = ","), ")"))

# 4. Indices for D3
links$IDsource <- match(links$source, node_totals$name) - 1
links$IDtarget <- match(links$target, node_totals$name) - 1

# 5. Render D3 Interactive Sankey
sankeyNetwork(Links = links, Nodes = node_totals, Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "label", fontSize = 14, nodeWidth = 40, 
              nodePadding = 20, sinksRight = FALSE, 
              colourScale = JS("d3.scaleOrdinal().range(['#1c2541', '#0077b6', '#00b4d8', '#90e0ef']);"))
```

### 5.3 Predictive Modelling (Risk Simulator)

The core of our Visual Analytics solution is transitioning from reactive data to proactive forecasting. We built a baseline **Multiple Linear/Logistic Regression Model** to predict `churn_probability` based on user behavior metrics. This mathematical model powers the "Risk Simulator" in our Shiny App, allowing stakeholders to use slider inputs to change these variables and instantly recalculate the predicted risk.

```{r}
# Build a baseline predictive model for Churn Probability
predictive_model <- lm(churn_probability ~ support_tickets_count + tx_count + satisfaction_score, data = df)

# Display the model summary to confirm statistical significance of variables
summary(predictive_model)
```
