---
title: "Project Minutes of Meeting"
author: "Team 15 - LIN XIAN, ZUO SU HUI"
date: today
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 2
---

## Scope of Work (SoW)

The scope of the **FinRetain** project encompasses the end-to-end development of an interactive R Shiny application designed to analyze, predict, and prescribe solutions for customer churn.

### **In-Scope:**

-   **Data Pipeline:** Cleaning and wrangling raw demographic and transactional data into a unified structure (`app_data.rds`).
-   **Analytics Engine:** Development of Descriptive (EDA), Diagnostic (CDA), Predictive (Survival/K-Means), and Prescriptive (Risk Simulator) modules.
-   **UI/UX Development:** Storyboarding, wireframing, and deploying a front-end Shiny dashboard for non-technical stakeholders.
-   **Documentation:** User guide, storyboard documentation, and final project poster.

### **Out-of-Scope:**

-   Live database integration (the app will use a static snapshot of data for this iteration).
-   User authentication and role-based access control (RBAC).

## Project Timeline

Below is the official record of team meetings, key decisions, and technical troubleshooting sessions conducted during the development of the FinRetain application.

------------------------------------------------------------------------

## Meeting 1: Project Kickoff & Data Strategy

**Date:** 21/02/2026\
**Time:** 16:00 - 18:00\
**Location:** Campus Library\
**Attendees:** LIN XIANWEI, SU HUI

### 🎯 Agenda

1.  Define project scope and core analytical modules.
2.  Discuss data handling for the large `transactions_data` file.
3.  Assign initial roles.

### 🗣️ Key Discussions & Decisions

-   **Data Pipeline:** The team identified that the transaction dataset is too large to process live on a free `shinyapps.io` server.
-   ***Decision:*** We will build an offline data pipeline to aggregate the transaction types and join them with the customer demographics, saving the output as a highly compressed `app_data.rds` file for lightning-fast app loading.
-   **Core Modules:** Agreed to build an EDA Demographic tab, a CDA Behavioral tab, and a Churn/CLV Risk Simulator using Linear Regression models.

### ✅ Action Items

| Task | Assignee | Deadline | Status |
|------------------|------------------|------------------|------------------|
| Create offline data wrangling script (`dplyr`, `tidyr`) | LIN XIANWEI | 21/02/2026 | Completed |
| Build base UI shell using `bslib` | LIN XIANWEI, SU HUI | 25/02/2026 | Completed |
| Train `lm` predictive models for Simulator | LIN XIANWEI, SU HUI | 27/02/2026 | Completed |

------------------------------------------------------------------------

## Meeting 2: Advanced Analytics Integration

**Date:** 01/03/2026\
**Time:** 15:00 - 18:00\
**Location:** Via Zoom\
**Attendees:** LIN XIANWEI, SU HUI

### 🎯 Agenda

1.  Review the base Shiny app.
2.  Brainstorm unique insights to secure high marks for novelty and methodology.

### 🗣️ Key Discussions & Decisions

-   **Novelty Requirement:** Standard bar charts are insufficient for top marks. We need advanced financial analytics.
-   ***Decision 1 (Survival Analysis):*** We will implement Kaplan-Meier survival curves using the `survival` package to predict the exact tenure month customers are most likely to churn.
-   ***Decision 2 (RFM Segmentation):*** We will calculate Recency, Frequency, and Monetary scores on the fly and visualize them using a `plotly` Treemap.
-   ***Decision 3 (Cash Flow):*** We will implement a `networkD3` Sankey diagram to visualize macro fund flows.

### ✅ Action Items

| Task | Assignee | Deadline | Status |
|------------------|------------------|------------------|------------------|
| Write dynamic RFM calculation code in `app.R` | LIN XIANWEI, SU HUI | 03/03/2026 | Completed |
| Build Sankey Diagram logic | LIN XIANWEI, SU HUI | 09/03/2026 | Completed |
| Implement Kaplan-Meier plots | LIN XIANWEI, SU HUI | 14/03/2026 | Completed |

------------------------------------------------------------------------

## Meeting 3: Deployment & Bug Squashing

**Date:** 15/03/2026\
**Time:** 14:00 - 17:00\
**Location:** Via Zoom\
**Attendees:** LIN XIANWEI, SU HUI

### 🎯 Agenda

1.  Deploy the unified `app.R` to shinyapps.io.
2.  Troubleshoot server crashes and UI bugs.

### 🗣️ Key Discussions & Decisions

-   **Issue 1: Server Crash on Deployment (Exit Status 1).** The server crashed because it was looking for the raw CSV files which were excluded from the deployment folder.
    -   ***Fix:*** Stripped the offline data pipeline from the deployment version of `app.R` and loaded `app_data.rds` directly.
-   **Issue 2: Survival Plot Error (`object of type 'symbol' is not subsettable`).** The `ggsurvplot` function failed in the reactive Shiny environment.
    -   ***Fix:*** Replaced `survfit()` with `surv_fit()` and explicitly wrapped the plot in a `print()` command.
-   **Issue 3: Sankey Diagram Visual Imbalance.** The left side of the Sankey diagram was visually squashed because incoming funds were much smaller than outgoing volume.
    -   ***Fix:*** Decoupled the visual drawing weights from the actual dollar amounts. Forced balanced proportions (100% vs 100%) while formatting the text labels to still display the actual, accurate currency amounts (`$X,XXX,XXX`).

------------------------------------------------------------------------

## Project Gantt Chart

The following Gantt chart visualizes our 12-week development sprint, color-coded by the phases of the Data Science Lifecycle.

```{r}
#| label: fig-gantt
#| fig-cap: "FinRetain Project Development Gantt Chart (Feb 21 - Apr 4)"
#| fig-width: 10
#| fig-height: 6

# Load required libraries
library(ggplot2)
library(dplyr)
library(lubridate)
library(forcats)

# 1. Create the Project Dataframe
project_tasks <- data.frame(
  Phase = c(rep("1. Initiation", 2), 
            rep("2. Data Prep", 1), 
            rep("3. Analytics", 4), 
            rep("4. Development", 3), 
            rep("5. Deployment", 2)),
  Task = c("Project Proposal & Methodology", 
           "UI/UX Storyboards & Wireframes",
           "Data Cleaning & RFM Engineering",
           "Module 1 & 2: EDA & CDA Scripting", 
           "Module 3: K-Means Clustering", 
           "Module 4: Survival Analysis", 
           "Module 5: Sankey Flow Logic",
           "Front-end UI Setup (bslib)", 
           "Integrating Analytics (apps.R)", 
           "Module 6: Risk Simulator",
           "QA Testing & Bug Fixing", 
           "Final Poster & Town Hall Prep"),
  Owner = c("LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI", "ZUO SU HUI", 
            "LIN XIANWEI", "LIN XIANWEI", "ZUO SU HUI", "ZUO SU HUI", 
            "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI & ZUO SU HUI", "LIN XIANWEI & ZUO SU HUI"),
  
  # Compressed timeline from Feb 21 to April 4 (with overlapping parallel tasks)
  Start = as.Date(c("2026-02-21", "2026-02-24", "2026-03-01", "2026-03-05", 
                    "2026-03-10", "2026-03-14", "2026-03-18", "2026-03-20", 
                    "2026-03-24", "2026-03-26", "2026-03-30", "2026-04-01")),
  End = as.Date(c("2026-02-28", "2026-03-02", "2026-03-08", "2026-03-12", 
                  "2026-03-17", "2026-03-20", "2026-03-25", "2026-03-27", 
                  "2026-03-31", "2026-04-02", "2026-04-03", "2026-04-04"))
)

# 2. Reorder tasks so the first task appears at the top of the chart
project_tasks <- project_tasks %>%
  mutate(Task = fct_reorder(Task, desc(Start)))

# 3. Generate the Gantt Chart using ggplot2
ggplot(project_tasks, aes(x = Start, xend = End, y = Task, yend = Task, color = Phase)) +
  geom_segment(linewidth = 8, lineend = "round") +
  geom_text(aes(x = Start, y = Task, label = Owner), 
            vjust = -1.5, size = 3, color = "darkgray", hjust = 0) +
  scale_color_brewer(palette = "Set1") +
  # Changed to show specific dates like "Feb 21", "Mar 07"
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(
    title = "FinRetain Project Roadmap",
    subtitle = "Task duration and ownership overview (Feb 21 - Apr 4)",
    x = "Timeline",
    y = NULL,
    color = "Project Phase"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray95", linetype = "dashed"),
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "bottom"
  )
```

------------------------------------------------------------------------

## Task Matrix & Peer Review Assignments

To ensure code quality and analytical rigor, every major task is assigned a specific "Owner" (the doer) and a "Reviewer" (the moderator).

```{r}
## 2. Project Gantt Chart

#| label: fig-timeline
#| fig-cap: "FinRetain Project Timeline and Milestones (Feb 21 - Apr 5)"
#| fig-width: 15
#| fig-height: 10

# Load required packages
pacman::p_load(vistime, ggplot2)

# Define project data using the csv text format with the Spring 2026 timeline
data <- read.csv(text="event,group,start,end,color
Project Proposal,1. Initiation,2026-02-21,2026-03-15,#a5d6a7
UI/UX Storyboards,1. Initiation,2026-02-21,2026-03-10,#a5d6a7
Data Cleaning & RFM,2. Data Prep,2026-03-01,2026-03-16,#81d4fa
EDA & CDA Modules,3. Analytics,2026-03-10,2026-03-24,#ce93d8
K-Means & Survival,3. Analytics,2026-03-15,2026-03-28,#ce93d8
Sankey Flow Diagram,3. Analytics,2026-03-20,2026-03-31,#ce93d8
R Shiny Dashboard,4. Development,2026-03-20,2026-04-03,#ffb74d
Risk Simulator Engine,4. Development,2026-03-25,2026-04-03,#ffb74d
Poster-Minutes-Guide,5. Deployment,2026-03-25,2026-04-01,#DD4B39
Presentation,6. Deployment,2026-04-01,2026-04-05,#DD4B39")

# Set the exact milestone deadlines
proposal_deadline <- as.Date("2026-03-15") 
poster_deadline <- as.Date("2026-04-01")
final_deadline <- as.Date("2026-04-05")

# Generate the timeline chart
p <- gg_vistime(data, title = "FinRetain Project Timeline (Spring 2026)") 

# Add vertical lines for the critical milestone deadlines
p + 
  geom_vline(xintercept = as.numeric(as.POSIXct(proposal_deadline)), color = "lightgreen", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = as.numeric(as.POSIXct(poster_deadline)), color = "orange", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = as.numeric(as.POSIXct(final_deadline)), color = "darkred", linewidth = 1.2, linetype = "solid") +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)
  )

#| label: tbl-assignments
#| tbl-cap: "Detailed Task Ownership and Moderation Roster"

pacman::p_load(knitr, kableExtra, dplyr)

# Create a dataframe for the Task Ownership Matrix with specific team members
task_matrix <- data.frame(
  Phase = c("1. Initiation", "2. Data Prep", "3. Analytics", "3. Analytics", "4. Development", "4. Development", "5. Deployment", "5. Deployment"),
  Task = c("Proposal & Storyboards", "Data Cleaning & RFM", "EDA & CDA Scripts", "K-Means & Survival", "Shiny App Setup", "Risk Simulator Engine", "Project Poster", "User Guide & Final App"),
  
  # Logically splitting the work between LIN XIANWEI and ZUO SU HUI
  Owner = c("LIN XIANWEI", "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI & ZUO SU HUI"),
  
  # Cross-reviewing each other's work
  Reviewer = c("ZUO SU HUI", "ZUO SU HUI", "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI", "ZUO SU HUI", "LIN XIANWEI", "Course Instructor")
)

# Format the table
task_matrix %>%
  kable(align = "l") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "left") %>%
  column_spec(1, bold = TRUE)
```

*Prepared by: TEAM 15 - LIN XIANWEI, ZUO SU HUI*
